---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Log-Concave Density Estimation Using the Score Matching Loss Function

`LogConcaveDESM` is an \code{R} package to compute and
visualize the log-concave score matching derocnsity estimate over a bounded domain.
It also contains functions to plot the first and second derivatives of the log-density estimate.
Furthermore, functions to compute various distances between two probability distributions
are provided to assess the quality of the density estimate.

## Installation

The development version of the `LogConcaveDESM` can be installed from the GitHub using `devtools`: 

```{r installation}
# install.packages("devtools")
devtools::install_github("zhoucx1119/LogConcaveDESM")
```

## Load the Package 

```{r warning=FALSE}
library(LogConcaveDESM, warn.conflicts = FALSE)
library(patchwork)
library(ggplot2)
```

## Computing and Visualizing Unpenalized Log-concave Score Matching Density Estimate

We show how to use functions in `LogConcaveDESM` to compute and visualize the log-concave 
score matching density estimate over a bounded domain. 

We first simulate data from a truncated normal distribution over the interval $[-5, 5]$. 

```{r load_data}
set.seed(2021)
N <- 200
parent_mean <- 0
parent_sd <- 1
domain <- c(-5, 5)
t_normal <- truncated_normal(parent_mean, parent_sd, domain)
data <- t_normal$sampling(N)
```

We plot the histogram of the data below. 

```{r hist_data}
binwidth <- 2 * stats::IQR(data) / N ** (1/3)
ggplot() + geom_histogram(
  data = data.frame(x = data), 
  aes(x = x, y = ..density..), 
  binwidth = binwidth,
            color = "darkblue",
            fill = "lightblue") + 
  coord_cartesian(xlim = domain, ylim = c(-0.01, 0.5)) + 
  theme_bw()
```

We assume that the second derivative of the logarithm of the log-concave score matching 
density estimate is a piecewise linear function with knots at the data points. 
With no penalty, the resulting optimal second derivative values at data points can be computed as below: 

```{r opt}
result <- lcd_scorematching(
  data = data, 
  domain = domain, 
  penalty_param = 0, 
  verbose = TRUE
)
```

Then, the second derivative of the logarithm of the log-concave score matching density estimate 
at arbitrary points can be computed using the `evaluate_logden_deriv2` function, and 
can be visualized using the `plot_logden_deriv2` function. 

```{r deriv2}
logden_deriv2_vals <- evaluate_logden_deriv2(
  scorematching_logconcave = result, 
  newx = seq(domain[1], domain[2], length.out = 11)
)
logden_deriv2_vals

plot_deriv2 <- plot_logden_deriv2(
  scorematching_logconcave = result, 
  plot_domain = domain, 
  plot_points_cnt = 500)
plot_deriv2
```

Subsequently, the first derivative of the logarithm of the log-concave score matching density estimate 
at arbitrary points can be computed using the `evaluate_logden_deriv1` function, and 
can be visualized using the `plot_logden_deriv1` function. 

```{r deriv1}
logden_deriv1_vals <- evaluate_logden_deriv1(
  scorematching_logconcave = result, 
  newx = seq(domain[1], domain[2], length.out = 11)
)
logden_deriv1_vals

plot_deriv1 <- plot_logden_deriv1(
  scorematching_logconcave = result, 
  plot_domain = domain, 
  plot_points_cnt = 500)
plot_deriv1
```

Then, the (un-normalized) log-density estimate at arbitrary point can be computed 
using the `evaluate_logdensity` function, and can be visualized using the `plot_logdensity` function. 

```{r logden}
logden_vals <- evaluate_logdensity(
  scorematching_logconcave = result, 
  newx = seq(domain[1], domain[2], length.out = 11)
)
logden_vals

plot_logden <- plot_logdensity(
  scorematching_logconcave = result, 
  plot_domain = domain, 
  plot_points_cnt = 500)
plot_logden
```

Finally, to evaluate and visualize the density estimate itself, we use `evaluate_density` 
and `plot_density` functions, respectively. 

```{r den}
den_vals <- evaluate_density(
  scorematching_logconcave = result, 
  newx = seq(domain[1], domain[2], length.out = 11)
)
den_vals

plot_den <- plot_density(
  scorematching_logconcave = result,
  plot_domain = domain,
  plot_points_cnt = 500, 
  plot_hist = TRUE
  )
plot_den
```

## Computing and Visualizing Penalized Log-concave Score Matching Density Estimate

As we can see above, the un-penalized log-concave score matching density estimate is 
too concentrated at the place where data are abundant. To remedy this, we consider 
penalized log-concave score matching density estimate, where the optimal penalty parameter 
is chosen using the `cv_optimal_density_estimate` function. 

```{r cv}
lambda_cand <- exp(seq(-5, 1, by = 0.5))
opt_den <- cv_optimal_density_estimate(
  data = data, 
  domain = domain, 
  penalty_param_candidates = lambda_cand, 
  fold_number = 5
)
```

The resulting density estimate is the following. 

```{r plot-opt}
plot_opt_den <- plot_density(
  scorematching_logconcave = opt_den,
  plot_domain = domain,
  plot_points_cnt = 500, 
  plot_hist = TRUE
  )
plot_opt_den
```

We could assess the quality of this density estimate using various metrics 
between two probability distributions, for example, the Kullback-Leibler divergence, 
the Hyvarinen divergence, the $L1$ distance, and the Hellinger distance. 

```{r metrics}
kl <- kl_div(
  true_density = t_normal, 
  density_estimate = opt_den) 

hyva <- hyvarinen_div(
  true_density = t_normal, 
  density_estimate = opt_den) 

l1 <- L1_dist(
  true_density = t_normal, 
  density_estimate = opt_den) 

he <- hellinger_dist(
  true_density = t_normal, 
  density_estimate = opt_den) 

metric_table <- cbind(
  Metrics = c("Kullback-Leibler Divergence", "Hyvarinen divergence", 
              "L1 Distance", "Hellinger Distance"), 
  Values = round(c(kl, hyva, l1, he), 5)
)

knitr::kable(metric_table)
```
